---
title: '[Optional] R Basics 1'
author: Christopher Gandrud
date: '2020-04-26'
slug: optional-r-basics-1
categories:
  - R
tags: [R]
Categories: []
Description: ''
Tags: [R]
runtime: shiny_prerendered
---

```{r setup-basics, include=FALSE}

```

# Welcome 

Welcome to to the course! This is the first of 2 sets of exercises introducing you to R, if you are not that familiar with it, or you would like to practice R basics before getting started.

# Getting Started with R and RStudio

To complete the exercises in this course, you will need to access RStudio. 

You can do this in the cloud on the course's **RStudio cloud** site:

- <https://rstudio.cloud/project/1140732>

Or download and **install** R and RStudio on your computer:

1. Install R: <https://cran.r-project.org/>

2. Install RStudio (free version): <https://rstudio.com/products/rstudio/download/>

# Install packages

In base R, use the `install.packages()` function to install the add-on packages you'll need to gather and manage data:

```{r eval=FALSE}
install.packages("tidyverse")
```

To use the packages, you need to theme with the `library()` function:

```{r eval=FALSE}
library("tidyverse")
```

Personally, I prefer (and will use for the rest of the course) `xfun::pkg_attach2()`. It installs packages if they aren't present and loads them all in one call.

```{r eval=FALSE}
xfun::pkg_attach2("tidyverse")
```

Note you will need to install the `xfun` package (`install.packages("xfun")`) if you haven't already.

# Basic concepts

- 